Future challenges for Research Software Engineering

### 2025-05-09 future_challenges_for_RSEng.tex
Add ideas GC

The digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Education, and changes in the educational system}
FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.

\subsection{Impact of AI and automation}
Is it better than cheap PhDs?
FG-NOTE: AI will change the interaction with researchers at a fundamental level. Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.

@@

MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?

@@

@@

The workshop participants found that in earlier learning stages,
details of systems may be intentionally hidden,
fearing a diminishing capability to handle technical complexity in research environments.
As a result, RSEs invest more effort in making individual research software components
less complex with better user experience.
This made participants worry about growing mismatch of peoples' expectations.
Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
No concrete examples were given on which complexities are now hidden behind abstractions.

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible,
allowing the scientific community to address questions not previously in grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
but it is also important to learn how to abstract from details,
by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

@@

>>>>

The digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\begin{FramedParagraphWithFootnotes}
GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernization, cleanup, and modularization,
which can pay off in faster and larger development in the future.
This is not possible by short-term funded solo RSEs.
\end{FramedParagraphWithFootnotes}

\subsection{Education, and changes in the educational system}

\begin{FramedParagraphWithFootnotes}
FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.
\end{FramedParagraphWithFootnotes}

\begin{FramedParagraphWithFootnotes}
GC-NOTE: Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.

GC-NOTE: By that point, we have an RSE Master's.
\end{FramedParagraphWithFootnotes}

\subsection{Impact of AI and automation}
Is it better than cheap PhDs?

\begin{FramedParagraphWithFootnotes}
FG-NOTE: AI will change the interaction with researchers at a fundamental level. Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.
\end{FramedParagraphWithFootnotes}

\begin{FramedParagraphWithFootnotes}
GC-NOTE: As we continue and expand documenting our work in the open and making our code available,
there will be more structured material to train GenAI on.

However, as people get quicker answers from GenAI, the community resources (Stack Overflow, Wikipedia) get poorer.
\end{FramedParagraphWithFootnotes}

@@

MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts? (GC: Where "product/artifacts", maybe also "publications"?)

@@

\begin{FramedParagraphWithFootnotes}
  GC-NOTE: New section on "New challenges":

  \begin{itemize}
    \item Networking of embedded RSEs and RSE units
    \item Everyone needs infrastructure: does everyone need to host their own infrastructure?
    \item Everyone hosts their infrastructure, but then everyone stays in silos? We need federated infrastructure.
  \end{itemize}
\end{FramedParagraphWithFootnotes}

@@

The workshop participants found that in earlier learning stages,
details may be intentionally hidden,
fearing a diminishing capability to handle technical complexity in research environments.
As a result, RSEs invest more effort in making individual research software components
less complex with better user experience.
This made participants worry about growing mismatch of peoples' expectations.
Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
No concrete examples were given on which complexities are now hidden behind abstractions.

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible,
allowing the scientific community to address questions not previously in grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
but it is also important to learn how to abstract from details
by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

GC-NOTE: Shall we add another story box here?

@@

GC-NOTE: RSE develop together with and for scientists.
Better software leads to better research,
better research leads to more better software.
